---
title: "OSCAR benchmarking for various p, n and optimizer settings"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{OSCAR benchmarking for various p, n and optimizer settings}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
\newcommand{\vertiii}[1]{{\left\vert\kern-0.25ex\left\vert\kern-0.25ex\left\vert #1 
    \right\vert\kern-0.25ex\right\vert\kern-0.25ex\right\vert}}
\newcommand{\bbeta}{\boldsymbol{\beta}}      # vector \beta bolded
\newcommand{\bg}{{\boldsymbol{g}}}           # vector g bolded
\newcommand{\bW}{{\boldsymbol{W}}}           # vector W bolded
\newcommand{\x}{{\boldsymbol{x}}}            # vector x bolded

```{r, echo=FALSE}
## Load necessary dependencies for running the benchmarks
# For OSCAR methodology; 
library(oscar)
# For 'omics data from prostate cancer (https://github.com/Syksy/curatedPCaData):
library(curatedPCaData)
# Surv
library(survival)
```

# OSCAR benchmarking

Prepare TCGA transcriptomics:

```{r tcga}
Y1 <- Surv(event=colData(mae_tcga)[["disease_specific_recurrence_status"]], time=colData(mae_tcga)[["days_to_disease_specific_recurrence"]])
X1_samps <- colData(mae_tcga)$sample_name[-which(is.na(Y1))]
Y1 <- Y1[-which(is.na(Y1))]
Y1 <- Y1[-which(!X1_samps %in% colnames(mae_tcga[["gex.rsem.log"]]))]
X1_samps <- X1_samps[-which(!X1_samps %in% colnames(mae_tcga[["gex.rsem.log"]]))]

length(X1_samps)
length(Y1)

X1 <- mae_tcga[["gex.rsem.log"]][,X1_samps]
# Remove genes with redundant expression, e.g. 0 across all samples
X1 <- X1[which(apply(X1, MARGIN=1, FUN=function(x) { !all(x==unique(x)[1]) })),]
```

Run benchmarks in TCGA

```{r}
# Run TCGA by default
benchf <- function(p, kmax, percentage, solver, X = t(X1), Y=Y1, iter = 1){
	t0 <- Sys.time()
	if(!missing(p)){
		fit <- oscar(x=X[,sample(1:ncol(X), p)], y=Y, family="cox", kmax=kmax, in_selection=3, percentage=percentage, solver=solver)
	}else{
		fit <- oscar(x=X, y=Y, family="cox", kmax=kmax, in_selection=3, percentage=percentage, solver=solver)
	}
	if(missing(p)){
		p <- ncol(X)
	}
	t1 <- Sys.time()
	dif <- t1 - t0
	c("Time diff" = as.numeric(t1 - t0), "Time unit" = attr(dif, "units"), "kmax" = kmax, "percentage" = percentage, "solver" = ifelse(solver=="LMBM", 2, 1), "f(kmax)" = fit@fperk[fit@kmax], "p" = p, "cindex" = fit@goodness[fit@kmax], iter = as.numeric(iter))	
}


set.seed(1)
benches <- benchf(kmax = 2, percentage = 0.01, solver = "LMBM")
benches

for(solv in "LMBM"){
	for(perc in c(0.00001, 0.0001, 0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
	#for(perc in c(0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
		for(k in c(2:5, 7, 10, 20, 30, 50)){
			bench <- benchf(kmax = k, percentage = perc, solver = solv, iter = 1)
			benches <- rbind(benches, bench)
			save.image("benchmarks_type1_tcga.RData")
		}
	}
}

set.seed(1)
benches2 <- benchf(kmax = 50, percentage = 0.01, p = 50, solver = "LMBM")
benches2

for(solv in "LMBM"){
	for(iteration in 1:50){
		for(perc in c(0.00001, 0.0001, 0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
		#for(perc in c(0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
			for(k in c(50)){
				for(p in c(50, 100, 200, 500, 1000, 2000, 5000, 10000)){
					set.seed(iteration)
					bench2 <- benchf(kmax = k, percentage = perc, solver = solv, p = p, iter = iteration)
					benches2 <- rbind(benches2, bench2)
					save.image("benchmarks_type2_tcga.RData")
				}
			}
		}
	}
}

deparse_bench <- function(x){
	# Time in minutes
	t <- apply(x, MARGIN=1, FUN=function(df){
		if(df["Time unit"]=="mins"){
			as.numeric(df["Time diff"])
		}else if(df["Time unit"]=="hours"){
			as.numeric(df["Time diff"])*60
		}else if(df["Time unit"]=="secs"){
			as.numeric(df["Time diff"])/60
		}
	})
	# 6th column is the f(kmax)
	data.frame(t_mins = t, k = as.numeric(x[,"kmax"]), perc = as.numeric(x[,"percentage"]), fkmax = as.numeric(x[,6]), p = as.integer(x[,"p"]), cindex = as.numeric(x[,"cindex.concordance"]), iter=as.integer(x[,"iter"]))
}
plot_timeperc <- function(df, title=""){
	dat <- by(df, INDICES=df$per, FUN=function(x){
		data.frame(x=x$k, y=x$t_mins)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$k), ylim=extendrange(df$t_mins))
	box(); axis(1); axis(2)
	title(xlab="k", ylab="Time (mins)", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]
		print(df)
		points(x=df$x, y=df$y, type="l", col=i)
		points(x=df$x, y=df$y, pch=16, col=i)	
	})
	legend("topleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}
plot_fvalperc <- function(df, title=""){
	dat <- by(df, INDICES=df$per, FUN=function(x){
		data.frame(x=x$k, y=x$fkmax)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$k), ylim=extendrange(df$fkmax))
	box(); axis(1); axis(2)
	title(xlab="k", ylab="f(k), -log likelihood", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]
		print(df)
		points(x=df$x, y=df$y, type="l", col=i)
		points(x=df$x, y=df$y, pch=16, col=i)	
	})
	#legend("bottomleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}
plot_cindexperc <- function(df, title=""){
	dat <- by(df, INDICES=df$per, FUN=function(x){
		data.frame(x=x$k, y=x$cindex)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$k), ylim=extendrange(df$cindex))
	box(); axis(1); axis(2)
	title(xlab="k", ylab="Concordance index", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]
		print(df)
		points(x=df$x, y=df$y, type="l", col=i)
		points(x=df$x, y=df$y, pch=16, col=i)	
	})
	#legend("bottomleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}
plot_nperc <- function(df, title=""){
	dat <- by(df, INDICES=df$perc, FUN=function(x){
		data.frame(x=x$p, y=x$t_mins)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$p), ylim=extendrange(df$t_mins))
	box(); axis(1); axis(2)
	title(xlab="dimensionality (p)", ylab="Time (mins)", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]
		print(df)
		#points(x=df$x, y=df$y, type="l", col=i)
		points(x=df$x, y=df$y, pch=16, col=i)	
	})
	#legend("topleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}
plot_timeiterperc <- function(df, title=""){
	dat <- by(df, INDICES=df$perc, FUN=function(x){
		data.frame(x=x$p, y=x$t_mins, iter=x$iter)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$p), ylim=extendrange(df$t_mins))
	box(); axis(1); axis(2)
	title(xlab="dimensionality (p)", ylab="Time (mins)", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]

		#print(df)

		df <- by(df, INDICES=df$iter, FUN=function(y){
			data.frame(x=y$x, y=y$y)
		})

		lapply(df, FUN=function(x){
			points(x=x$x, y=x$y, lwd=1, col=i, type="l")	
			points(x=x$x, y=x$y, pch=16, col=i)	
		})
	})
	#legend("topleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}
plot_fvaliterperc <- function(df, title=""){
	dat <- by(df, INDICES=df$perc, FUN=function(x){
		data.frame(x=x$p, y=x$fkmax, iter=x$iter)
	})
	percs <- as.numeric(names(dat))
	
	plot.new()
	plot.window(xlim=extendrange(df$p), ylim=extendrange(df$fkmax))
	box(); axis(1); axis(2)
	title(xlab="dimensionality (p)", ylab="f(k), -log likelihood", main=title)
	
	lapply(1:length(percs), FUN=function(i){
		df <- dat[[i]]

		#print(df)

		df <- by(df, INDICES=df$iter, FUN=function(y){
			data.frame(x=y$x, y=y$y)
		})

		lapply(df, FUN=function(x){
			points(x=x$x, y=x$y, lwd=1, col=i, type="l")	
			points(x=x$x, y=x$y, pch=16, col=i)	
		})
	})
	#legend("topleft", lwd=1, pch=16, col=1:length(dat), legend=paste("Percentage = ", names(dat)))
	dat
}

load("benchmarks_type1.RData")
load("benchmarks_type2.RData")

## Plot results from TCGA

datbench <- deparse_bench(benches)
datbench2 <- deparse_bench(benches2)

#par(mfrow=c(2,3))
par(mfrow=c(2,2))
# Upper row
dattime_a1 <- plot_timeperc(datbench, title="TCGA full (p=19,358, n=413)")
dattime_a2 <- plot_fvalperc(datbench, title="TCGA full (p=19,358, n=413)")
dattime_a3 <- plot_cindexperc(datbench, title="TCGA full (p=19,358, n=413)")
# Lower row
#dattime_b1 <- plot_nperc(datbench2, title="TCGA subsampled p (n=413, kmax=50)")
dattime_b2 <- plot_timeiterperc(datbench2, title="TCGA subsampled p (n=413, kmax=50)")

```


Process MSKCC transcriptome:

```

X2 <- mae_taylor[["gex.rma"]]
Y2 <- Surv(event=colData(mae_taylor)[["disease_specific_recurrence_status"]], time=colData(mae_taylor)[["days_to_disease_specific_recurrence"]])
names(Y2) <- colData(mae_taylor)[,"sample_name"]
X2_samps <- colData(mae_taylor)$sample_name[-which(is.na(Y2))]
Y2 <- Y2[-which(is.na(Y2))]
Y2 <- Y2[-which(!X2_samps %in% colnames(mae_taylor[["gex.rma"]]))]
X2_samps <- X2_samps[-which(!X2_samps %in% colnames(mae_taylor[["gex.rma"]]))]
X2 <- mae_taylor[["gex.rma"]][,X2_samps]
# Remove genes with redundant expression, e.g. 0 across all samples
X2 <- X2[which(apply(X2, MARGIN=1, FUN=function(x) { !all(x==unique(x)[1]) })),]



benchf <- function(p, kmax, percentage, solver, X = t(X2), Y=Y2, iter = 1){
	t0 <- Sys.time()
	if(!missing(p)){
		fit <- oscar(x=X[,sample(1:ncol(X), p)], y=Y, family="cox", kmax=kmax, in_selection=3, percentage=percentage, solver=solver)
	}else{
		fit <- oscar(x=X, y=Y, family="cox", kmax=kmax, in_selection=3, percentage=percentage, solver=solver)
	}
	if(missing(p)){
		p <- ncol(X)
	}
	t1 <- Sys.time()
	dif <- t1 - t0
	c("Time diff" = as.numeric(t1 - t0), "Time unit" = attr(dif, "units"), "kmax" = kmax, "percentage" = percentage, "solver" = ifelse(solver=="LMBM", 2, 1), "f(kmax)" = fit@fperk[fit@kmax], "p" = p, "cindex" = fit@goodness[fit@kmax], iter = as.numeric(iter))	
}

set.seed(1)
benches <- benchf(kmax = 2, percentage = 0.01, solver = "LMBM")
benches

for(solv in "LMBM"){
	for(perc in c(0.00001, 0.0001, 0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
	#for(perc in c(0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
		for(k in c(2:5, 7, 10, 20, 30, 50)){
			bench <- benchf(kmax = k, percentage = perc, solver = solv, iter = 1)
			benches <- rbind(benches, bench)
			save.image("benchmarks_type1_mskcc.RData")
		}
	}
}

set.seed(1)
benches2 <- benchf(kmax = 50, percentage = 0.01, p = 50, solver = "LMBM")
benches2

for(solv in "LMBM"){
	for(iteration in 1:50){
		for(perc in c(0.00001, 0.0001, 0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
		#for(perc in c(0.0002, 0.0005, 0.001, 0.005, 0.01, 0.02)){
			for(k in c(50)){
				for(p in c(50, 100, 200, 500, 1000, 2000, 5000, 10000)){
					set.seed(iteration)
					bench2 <- benchf(kmax = k, percentage = perc, solver = solv, p = p, iter = iteration)
					benches2 <- rbind(benches2, bench2)
					save.image("benchmarks_type2_mskcc.RData")
				}
			}
		}
	}
}

## Plot results from Taylor et al. (MSKCC)

load("benchmarks_type1_mskcc.RData")
load("benchmarks_type2_mskcc.RData")

datbench <- deparse_bench(benches)
datbench2 <- deparse_bench(benches2)

#par(mfrow=c(2,3))
par(mfrow=c(2,2))
# Upper row
dattime_a1 <- plot_timeperc(datbench, title="MSKCC full (p=17,410, n=140)")
dattime_a2 <- plot_fvalperc(datbench, title="MSKCC full (p=17,410, n=140)")
dattime_a3 <- plot_cindexperc(datbench, title="MSKCC full (p=17,410, n=140)")
# Lower row
#dattime_b1 <- plot_nperc(datbench2, title="MSKCC subsampled p (n=140, kmax=50)")
dattime_b2 <- plot_timeiterperc(datbench2, title="MSKCC subsampled p (n=140, kmax=50)")



```




# Session info

```{r session}
sessionInfo()
```

