---
title: "Example use of the OSCAR package"
author: "Teemu Daniel Laajala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example use of the OSCAR package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r oscar}
library(oscar)
```


# Example use of OSCAR

OSCAR (Optimal Subset CArdinality Regression) is an R-package delivering L0-quasinorm penalized regression models for Cox, logistic and gaussian model families. 
It comes with two different optimizers for the computationally demanding task, and they have been implemented in Fortran.
The R package handles all interaction with Fortran and provides many convenience and visualization functions to assist with L0-penalized regression modelling.

## Cox regression (simulated TYKS real-world PCa data)

A simulated real-world hospital cohort data with patient survival with prostate cancer (PCa) is provided with:
```{r exdata}
data(ex)
ex_X[1:10,1:10]
head(ex_Y)
```

For further details regarding this dataset, see reference [1].

One key feature in OSCAR in addition to its L0 norm penalization is its capability to handle variables as groups, or 'kits' as we call them. 
In many clinical applications this is often reality; for example, curating initial data for variables together can occur in cases such as:

* A series of laboratory tests (such as a generic blood panel) are run together
* A pre-made small gene panel is run, instead of transcriptome-wide analysis
* Highly correlated values are run together

In the example data from Turku University Hospital (TYKS), a series of clinical variables are run together. 
For example a standard blood panel includes red blood cell count, hematocrite, white blood cell count, etc. 
Additionally, single variables can be run separately; a very common single marker for PCa is Prostate-Specific Antigen (PSA).

The kit structure for such variables that can be obtained together are modelled using a kit indicator matrix:
```{r kits}
ex_K[1:7,1:7]
apply(ex_K, MARGIN=1, FUN=sum) # Row indicator sums
apply(ex_K, MARGIN=2, FUN=sum) # Column indicator sums
```

If the kit matrix is a diagonal indicator matrix with dimensions equal to the number of variables, each variable is L0-penalized without any grouping structure.

Further, these kits can be assigned a 'price'. This may be a price literally, or a descriptive value indicating how hard the variable is to obtain. 
With this information, a clinically feasible pareto-front can be constructed, that simultaneously captures modelling generalization as well as clinical applicability.

A cost vector can be provided for the fitting procedure, and a standardized arbitrary unit cost vector is provided in:

```{r cvec}
head(ex_c)
```

OSCAR comes with two different optimizers: 

* DBDC, double bundle DC optimization, which is computationally more demanding but in general results in better optima
* LMBM, which is faster to run, especially with higher count of variables

These are provided to the fitting procedure with ```solver=1``` and ```solver=2``` or with their abbreviations. They have been implemented in Fortran for computational efficiency.

A quick example run of OSCAR with LMBM together with the default kit structure for Cox regression:

```{r}
fit <- oscar(x=ex_X, y=ex_Y, k=ex_K, w=ex_c, family="cox", solver="LMBM")
fit
```

A pareto-optimal front provides information on potential optimal saddle points, where model generalization and clinical applicability pair nicely:

```{r pareto, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Pareto front for the clinical measurements and C-index."}
plot.pareto.costCI(costs=fit@w, cindexes=fit@goodness, nzeros=1:fit@kmax)
```

## Gaussian regression (swiss-data)

```{r eval=FALSE, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Model coefficients as a function of k."}

# Use example swiss-data for quickness
library(casso)
data(swiss)
fit <- oscar(x=swiss[,-1], y=swiss[,1], family="mse", print=0)

# Plot model coefficients as a function of cardinality k
plot(fit) 

```

### Goodness of fit vs. model coefficient cost

Model convergence in target function is important, while an another axis can be plotted simultaneously to provide supporting information:

```{r eval=TRUE, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Target function value as a function of k."}

visu(fit, y="target")

```

### Cross-validation and bootstrapping

Fit model object can be inspected based on visual diagnostics such as cross-validation and bootstrapping:

```{r eval=TRUE, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="10-fold cross-validation for optimally generalizable cardinality 'k'"}
# Set seed for reproducibility
set.seed(1)
# Perform 10-fold cross-validation to find out optimal k
cv <- cv.oscar(fit, fold=5, seed=123, solver=2)
# Visualize model generalization performance as a function of k
cv.visu(cv)

```

It appears that k-values around 5 or above present a shoulder-point saturate the model, with no further coefficients contributing to the model's generalization capability. 

Further, there is some uncertainty as to what order the coefficients should be non-zero, based on bootstrapping:

```{r eval=TRUE, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Bootstrapped coefficient trajectories"}
# Bootstrap original data 20 times (sampling with replacement and refitting)
bs <- bs.oscar(fit, bootstrap=20, seed=234, solver=2)
# Visualize trajectories of bootstrapped coefficients
bs.plot(fit=fit, bs=bs, nbins=20)

```

Based on the bootstrapping, PSA is a clear winner. Close second candidate for survival prediction is Alkaline Phosphatase (ALP). 
After these single markers the bootstrapping was quite prone to choosing separate kits between immunomarkers and standard bloodwork, with no clear consistent winner.

### Further useful commands

Useful commands for understanding the model and the casso object:

```{r eval=TRUE}

coef(fit, k=3) # All potential coefficients at cardinality k=3
feat(fit, k=3) # All features chosen at cardinality k=3
cost(fit, k=3) # Kit sum costs at various k cardinalities

```

# References

1: Laajala TD, Murtojarvi M, Virkki A, Aittokallio T. ePCR: an R-package for survival and time-to-event prediction in advanced prostate cancer, applied to real-world patient cohorts. Bioinformatics. 2018;34(22):3957-3959. doi:10.1093/bioinformatics/bty477

# Session info
```{r sesinfo}
sessionInfo()
```