---
title: "Example use of the CASSO package"
author: "Teemu Daniel Laajala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example use of the CASSO package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Example use of CASSO and fine-tuning parameters

```{r, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Model coefficients as a function of k."}

# Use example swiss-data for quickness
library(casso)
data(swiss)
fit <- casso(x=swiss[,-1], y=swiss[,1], family="mse", print=0)

# Plot model coefficients as a function of cardinality k
plot(fit) 

```

Model convergence in target function is important, while an another axis can be plotted simultaneously to provide supporting information:

```{r, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Target function value as a function of k."}

visu(fit, y="target")

```

Fit model object can be inspected based on visual diagnostics such as cross-validation and bootstrapping:

```{r, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="10-fold cross-validation for optimally generalizable cardinality 'k'"}

# Perform 10-fold cross-validation to find out optimal k
cv <- cv.casso(fit, fold=10, seed=0)
# Visualize model generalization performance as a function of k
cv.visu(cv)

```

Appears that k-values 3 or 4 saturate the model, with no further coefficients contributing to the model's generalization capability. Further, there is some uncertainty as to what order the coefficients should be non-zero, based on bootstrapping:

```{r, dpi=150, fig.width=7, fig.height=7, out.width="100%", fig.cap="Bootstrapped coefficient trajectories"}

# Bootstrap original data 25 times (sampling with replacement and refitting)
bs <- bs.casso(fit, bootstrap=25, seed=0)
# Visualize trajectories of bootstrapped coefficients
bs.visu(bs)

```

Useful commands for understanding the model and the casso object:

```{r}

coef(fit, k=3) # All potential coefficients at cardinality k=3
feat(fit, k=3) # All features chosen at cardinality k=3

```
